A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-21 main.tex
3.6 edits

\pb{I prefer ``synthetic'' phenotype to pseudophenotype}

@@

\subsubsection{Design matrix generation}

\pb{Let's not call it a ``design matrix''}

\pb{Also, we're generating $G$ matrices, correct?  And then obtaining $X$ from standardization?  Wouldn't hurt to remind people of that here}

@@

We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \anna{cite Vanderwas data PATRICK TO DO}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Semi-Empirical data, respectively. \pb{Don't start sentences with mathematical symbols} $p = 1000$ for all data types, $n = \{200, 400, 600, 800 \}$ for the 1D Linear Admixture and Independent Subpopulations data, and 50 causal SNPs were used for all data types.

\pb{Can't we just call the last one ``Empirical''?}

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered, which we refer to as ``fine'' and ``coarse''.  We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, representing a continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations (also known as founding populations) are of equal size. We note that the distinction between fine and coarse structure is an entirely different concept than the level of admixture in the simulated data. \anna{size of the subpopulations that determines fine/course - cut into 2 sentences. similarly for the 1d amixture...} For example, the discrete nonoverlapping subpopulations in the simulated Independent Subpopulations data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated 1D Linear Admixture data \pb{?????}. The Semi-Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ? MAKE THESE AND MAKE THEM PRETTY} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Semi-Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. \anna{Qualitatively, this is due to the fact that the Independent Subpopulations and Semi-Empirical data sets are characterized by more discrete, nonoverlapping genetic subpopulations, which more directly correspond to the distinct fixed environmental effects than do the continuous and blended subpopulations of the 1D Linear Admixture data. In other words, a greater proportion of the environmental effect is in the column space of the projection matrix of $\bX$ for the Semi-Empirical and Independent Subpopulations data than for the 1D Linear Admixture data. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the ratio \eqref{eqn:ratio} introduced in Section \ref{sec:sim_env_conf}. $||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2$, is equal to 0.98615, 0.64325, and 0.46412 for the Independent Subpopulations, Semi-Empirical, and 1D Linear Admixture data, respectively. As such there is a greater amount of confounding and estimation bias which the PC-Lasso and LMM-Lasso methods are able to correct for in the Independent Subpopulations, while there is less confounding which can be corrected for in the 1D Linear Admixture data.}

@@

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean TPR and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Semi-Empirical data sets than in the 1D Linear Admixture data. \anna{Again, this is due to the smaller proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ for the 1D Linear Admixture data, which results in less confounded SNP-phenotype associations to begin with.}

\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Semi-Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-Lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & Lasso & 0.23 (0.05) & 0.25 (0.05) & 0.29 (0.06) \\ 
& & PC-Lasso & 0.28 (0.05) & 0.36 (0.06) & 0.47 (0.06) \\ 
& & LMM-Lasso & 0.29 (0.05) & 0.39 (0.06) & 0.58 (0.06) \\ 
\bottomrule
\end{tabularx}
\caption{True positive rates, mean (SD), for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
\label{tab:var_sel}
\end{table}

>>>>

@@

\subsubsection{SNP matrix generation}

@@

We generated these synthetic phenotypes using three types of high-dimensional ($p > n$) $\bG$ matrices, which were then standardized to yield the corresponding $\bX$ matrix: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \anna{cite Vanderwas data PATRICK TO DO}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. For all data types $p = 1000$ with 50 causal SNPs. For the simulated 1D Linear Admixture and Independent Subpopulations data $n = \{200, 400, 600, 800 \}$, and $n = 588$ for the Empirical data.

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered, which we refer to as ``fine'' and ``coarse''.  We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, representing a continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations (also known as founding populations) are of equal size. We note that the distinction between fine and coarse structure is an entirely different concept than the level of admixture in the simulated data. \anna{size of the subpopulations that determines fine/course - cut into 2 sentences. similarly for the 1d amixture...} For example, the discrete nonoverlapping subpopulations in the simulated Independent Subpopulations data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated 1D Linear Admixture data \pb{?????}. The Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ? MAKE THESE AND MAKE THEM PRETTY} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for Lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), Lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the ordinary Lasso. This improvement is more substantial in the Empirical and Independent Subpopulations data than in the 1D Linear Admixture data. \anna{Qualitatively, this is due to the fact that the Independent Subpopulations and Empirical data sets are characterized by more discrete, nonoverlapping genetic subpopulations, which more directly correspond to the distinct fixed environmental effects than do the continuous and blended subpopulations of the 1D Linear Admixture data. In other words, a greater proportion of the environmental effect is in the column space of the projection matrix of $\bX$ for the Empirical and Independent Subpopulations data than for the 1D Linear Admixture data. We can quantify the proportion of the environmental component in the column space of $\boldsymbol{P}_{\bX}$ for each data type using the ratio \eqref{eqn:ratio} introduced in Section \ref{sec:sim_env_conf}. $||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2$, is equal to 0.98615, 0.64325, and 0.46412 for the Independent Subpopulations, Empirical, and 1D Linear Admixture data, respectively. As such there is a greater amount of confounding and estimation bias which the PC-Lasso and LMM-Lasso methods are able to correct for in the Independent Subpopulations, while there is less confounding which can be corrected for in the 1D Linear Admixture data.}

@@

In addition to MSE, these methods' performance in terms of variable selection accuracy were quantified using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Table \ref{tab:var_sel} displays the mean TPR and corresponding standard deviation in the same settings as those illustrated in Figure \ref{fig:mse}. The TPRs echo the MSE results in that Lasso and LMM-Lasso perform comparably in the absence of environmental confounding across all data types, while the addition of the 10 unpenalized PCs leads to a slight decrease in TPR in this setting for the 1D Linear Admixture and Independent Subpopulations data sets. In the presence of environmental confounding, both LMM-Lasso and PC-Lasso achieve substantial improvements in TPR over Lasso across data types. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data. \anna{Again, this is due to the smaller proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ for the 1D Linear Admixture data, which results in less confounded SNP-phenotype associations to begin with.}

\begin{table}[H]
\centering
\begin{tabularx}{400pt}{cclYYY}
\toprule
$\eta$ & $\xi$ & \multicolumn{1}{c}{Method} & 1D Linear Admixture & Independent Subpopulations & Empirical \\ 
\midrule
\multirow{6}{*}{0.5} & \multirow{3}{*}{0} & Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.37 (0.06) \\ 
& & PC-Lasso & 0.23 (0.05) & 0.21 (0.05) & 0.36 (0.06) \\ 
& & LMM-Lasso & 0.25 (0.05) & 0.23 (0.05) & 0.41 (0.06) \\
\cmidrule{2-6}
& \multirow{3}{*}{0.8} & Lasso & 0.23 (0.05) & 0.25 (0.05) & 0.29 (0.06) \\ 
& & PC-Lasso & 0.28 (0.05) & 0.36 (0.06) & 0.47 (0.06) \\ 
& & LMM-Lasso & 0.29 (0.05) & 0.39 (0.06) & 0.58 (0.06) \\ 
\bottomrule
\end{tabularx}
\caption{True positive rates, mean (SD), for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
\label{tab:var_sel}
\end{table}
